---
title: GLM Poisson and Negative Binomial (overdispersed Poisson)
author: Wesley, Ezequiel Marzinelli
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(shiny)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stargazer)
library(lattice)  
library(gridExtra)
library(magrittr)
library(xtable)
theme_set(theme_bw())
```

```{r load, echo=FALSE}
Benthos = read.table(file = "../data/infauna.txt", h = T, dec = ".")
#dec = '.'   means that the point is used for decimals. 
#Change to   dec = ","   if required.
```

## Basic Exploratory Analysis

```{r}
names(Benthos)
str(Benthos)
```

## Aim of the analysis:

To model Ampeliscidae as a function of the covariates:

* Period
* Fishing
* OrganicM
* Mud
* Silt
* Clay
with interaction between Fishing and OrganicM.

| Variable | Definition            |
| ---      | ---                   |
| Fishing  | Fishing vs no fishing |
| Period   | Three time periods    |

```{r Housekeeping}
#Converting Period and Fishing into factors
Benthos %<>% mutate(fPeriod = factor(Period), 
                    fFishing = factor(Fishing, 
                                      levels = c("no", "yes"), 
                                      labels = c("No Fishing", "Fishing")))
```

## BASIC DATA EXPLORATION:
1. Outliers: Y and X
2. Collinearity X
3. Relationships  Y vs X    but also interactions
4. Zero inflation: Make a frequency plot


### 1. Are there any outliers?
  * Outliers in the response variable?
  * #Outliers in the explanatory variables?

```{r}
Benthos                                         %>%
    select(Ampeliscidae, OrganicM, Silt, Clay, Mud) %>%
    gather(Var, Value)                              %>%
    ggplot()                           +
        geom_jitter(aes(factor(1), Value)) +
        facet_wrap(~Var)
```

Great because response var ranges till 15 (<20-25), Poisson may be good.

#Numer of zeros in the response variable

```{r, results='asis'}
table(Benthos$Ampeliscidae) %>% xtable
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(Ampeliscidae, geom="bar", data = Benthos)
```

Percentage  of zeros in data: `r 100 * sum(Benthos$Ampeliscidae == 0) / nrow(Benthos)`%.

### What about categorical covariates?

Do we have a reasonable number of observations per level of a categorical covariate?

```{r}
with(Benthos, table(fFishing)) #with the object "Benthos", give me a table
with(Benthos, table(fPeriod))
with(Benthos, table(fPeriod,fFishing)) 
#if I was interested in their interaction
#Nice, more or less balanced design
```

### 2. Check for Collinearity

```{r}
pairs(Benthos[, c("OrganicM", "Mud", "Silt", "Clay")])
```

```{r results='asis'}
#gives you a correlation matrix
cor(Benthos[, c("OrganicM", "Mud", "Silt", "Clay")]) %>% stargazer(type='html')
```

**Conclusion**: Collinearity is observed between Clay and Silt and Mud

```{r}
df = Benthos                                         %>%
    select(Ampeliscidae, OrganicM, Silt, Clay, Mud, fPeriod, fFishing) %>%
    gather(Var, Value, -fFishing, -fPeriod)                             
```

```{r}
df %>% ggplot(aes(x=fPeriod, y=Value)) +
    geom_boxplot()                     +
    geom_jitter()                      +
    facet_wrap(~Var, scales="free")    + 
    ggtitle("Period")
```

```{r}
df %>% 
    ggplot(aes(x=fFishing, y=Value)) +
    geom_boxplot()                   +
    geom_jitter()                    +
    facet_wrap(~Var, scales="free")  +
    ggtitle("Fishing")
```

```{r}
pPeriod = df %>% filter(Var == 'Silt') %>%
    ggplot(aes(x=fPeriod, y=Value))                           +
    geom_boxplot()+ 
    geom_jitter()
pFishing = df %>% filter(Var == 'Silt') %>%
    ggplot(aes(x=fFishing, y=Value))                           +
    geom_boxplot()+
    geom_jitter()
grid.arrange(pPeriod, pFishing, nrow=1, main="Silt")
```

```{r}
pPeriod = df %>% filter(Var == 'Clay') %>%
    ggplot(aes(x=fPeriod, y=Value))                           +
    geom_boxplot()+ 
    geom_jitter()
pFishing = df %>% filter(Var == 'Clay') %>%
    ggplot(aes(x=fFishing, y=Value))                           +
    geom_boxplot()+
    geom_jitter()
grid.arrange(pPeriod, pFishing, nrow=1, main="Clay")
```

#### Conclusion:

Either use mud or silt and clay. 
Fishing seems to be collinear with some of these.
If you use mud or clay you cannot use fishing as it is collinear!

### Check for Interactions

Based on biology we would expect that Ampeliscidae vs O_Material,
changes depending on the dredging effect (CT)!!!

```{r}
qplot(OrganicM, Ampeliscidae, data=Benthos) + facet_wrap(~fFishing) + 
  stat_smooth(method = "lm", formula = y ~ x, na.rm=T)
```
#There is indication for interaction!!

### Start modelling analysis

Fit following the model:
E(Ampeliscidae ) = mu = exp(alpha + A + B + A:B + C )

```{r}
M1 =  glm(Ampeliscidae ~ OrganicM * fFishing + fPeriod, 
          data = Benthos,
          family = poisson)
         
#Note this is exactly the same:
M1 <- glm(Ampeliscidae ~ OrganicM + fFishing +
                         OrganicM : fFishing +
                         fPeriod, 
          data = Benthos,
          family = poisson)
```

```
What is the model that we are fitting?
Ampeliscidae_i ~ P(mu_i)
E(Ampeliscidae_i) = mu_i)
and   var(Ampeliscidae_i) = mu_i

log(mu_i) = alpha + OrganicM_i + fFishing_i + 
            Period_i + OrganicM_i x fFishing_i
```

#### Check for overdispersion

**Pearson residuals** 
```
  (Y - E(y))          (Y - mu)
 --------------  = -----------------
  sqrt(var(Y))         sqrt(mu)
mu = exp(blah blah blah)
```

```{r}
E1 <- resid(M1, type = "pearson")
N  <- nrow(Benthos)
p  <- length(coef(M1))
#What is this?
sum(E1^2) / (N - p)
```

```{r results='asis'}
stargazer(M1, type='html')
```

```{r results='asis'}
drop1(M1, test = "Chi") %>% stargazer(type='html')
```

Just ok!

Instead of p-values summary statistic uses: (LRT) **Likelihood Ratio Test**

#### Suppose that the interaction is not significant:

```{r eval=FALSE}
Mtest <- glm(Ampeliscidae ~ OrganicM + fFishing  + fPeriod, 
          data = Benthos,
          family = poisson)
summary(Mtest)
drop1(Mtest, test =  "Chi")
```

### Model validation

Plot residuals vs fitted values
#Influential observations
Plot residuals vs each covariate (in the model, and not in the model)

```{r}
F1 <- fitted(M1) #command fitted gives already e^model
E1 <- resid(M1, type = "pearson")
plot(x = F1, 
     y = E1,
     xlab = "Fitted values",
     ylab = "Pearson residuals")
abline(h = 0, lty = 2)
```


#### Cook's distance: influential obs

There's a paper in the email which you can use (not usable with random effects)
```{r}
par(mfrow = c(1, 1))
plot(M1, which = 4)
#Plot Pearson residuals versus each covariate
plot(x = Benthos$OrganicM, 
     y = E1)
abline(h = 0, lty = 2)
#do the same for other covariates not in model 
plot(x = Benthos$Clay, 
     y = E1)
abline(h = 0, lty = 2)
#Plot3
plot(x = Benthos$Mud, 
     y = E1)
abline(h = 0, lty = 2)
#Plot4 
plot(x = Benthos$Silt, 
     y = E1)
abline(h = 0, lty = 2)
#Plot5
boxplot(E1 ~ fPeriod, data = Benthos) 
#Plot6
boxplot(E1 ~ fFishing, data = Benthos) 
```

### Model interpretation

```{r}
#Sketch fitted values  for the GLM Poisson model
M1 <- glm(Ampeliscidae ~ OrganicM * fFishing + fPeriod, 
          data = Benthos,
          family = poisson)
```

```{r results='asis', echo=FALSE}
stargazer(M1, type='html')
```

#### Qn. What is the model that we are fitting?

```
Ampeliscidae_i ~ Poisson(mu_i)
E(Ampeliscidae_i)   = mu_i
var(Ampeliscidae_i) = mu_i
         alpha + OrganicM + Fishing + Period + OrganicM:Fishing  
mu_i = e
```

#### Qn. What are the equations for each period/Fishing combination?
```
What is the equation for Period = 1 & Fishing = Non Fishing
         -0.1766 + 0.657 * OrganicM_i  
mu_i = e
```

##### Qn. What is the equation for Period = 2 & CT = Non-Fishing
```
         -0.1766 +  1.1919 + 0.657 * OrganicM_i  
mu_i = e
```

##### Qn. What is the equation for Period = 3 & CT = Non-Fishing
```
         -0.1766 +  1.4371 + 0.657 * OrganicM_i  
mu_i = e
```

##### Qn. What is the equation for Period = 1 & CT = Fishing
```
         -0.1766 + 1.3694 + (0.657 -1.1314) * OrganicM_i  
mu_i = e
```


##### Qn. What is the equation for Period = 2 & CT = Fishing
```
-0.1766 + 1.3694 + 1.19 + (0.657 -1.1314) * OrganicM_i  
#mu_i = e
```

```{r}
#1 Generate exploenatory tdata points
myData = Benthos %>% select(fPeriod, fFishing) %>% unique %$%
do.call(rbind,mapply(
    function(fFishing,fPeriod){
          data.frame(
              OrganicM = with(Benthos, seq(min(OrganicM),max(OrganicM),len=10)),
              fPeriod,
              fFishing
          )
    },
    fFishing = fFishing,
    fPeriod  = fPeriod,
    SIMPLIFY = FALSE))
myData %>% head
```

```{r}
#2 Predict dependant variable
myData %<>%
    group_by(fPeriod, fFishing) %>%
    do({
       data.frame(.,Ampeliscidae= predict(M1, newdata= ., type="response", se=TRUE))    
       #whats the difference between type=link, link doesnt transform for you
    })
```

```{r}
p0 = ggplot(Benthos)+
        geom_point(aes(OrganicM, Ampeliscidae, color = factor(fFishing), shape=fPeriod)) +
        labs(y = "Ampeliscidae",x = "Organic Matter")+scale_color_discrete("Fishing")
#All points
p0 +
geom_smooth(data = myData, 
                aes(y=Ampeliscidae.fit, x=OrganicM, 
                    ymin = Ampeliscidae.fit-Ampeliscidae.se.fit, ymax = Ampeliscidae.fit+Ampeliscidae.se.fit,
                    color=fFishing,
                    linetype=fPeriod
                    ), stat = "identity")
```

```{r echo=FALSE}
shinyApp(
       ui =  bootstrapPage(
                inputPanel(
                    selectInput(inputId = "period",  label = "Period",  choices = c(1,2,3), selected=1),
                    selectInput(inputId = "fishing", label = "Fishing", choices = c("Fishing", "No Fishing"), selected="Fishing")
                        ),
                mainPanel(
                    plotOutput("plot")
                )
            ),
        server = function(input, output){

        theDF = reactive ({myData %>% filter(fPeriod == input$period, fFishing == input$fishing)
            })

        title = renderText({
            sprintf("%s and %s", input$period, input$fishing)
        })
        output$plot= renderPlot(
        p0 +
                        geom_smooth(data = theDF(),
                            aes(y=Ampeliscidae.fit, x=OrganicM,
                                ymin = Ampeliscidae.fit-Ampeliscidae.se.fit, ymax = Ampeliscidae.fit+Ampeliscidae.se.fit), 
                                    stat = "identity")+ggtitle(title())
        )
        }, 
         options = list(height = 800)
)
```

Cirratulidae (Second Species)
====

Similar data exploration steps

```{r eval=FALSE, echo=FALSE}
#Home work: do the data exploration
dotchart(Benthos$Cirratulidae)

#Start analysis
#Fit the model E[Y] = mu = exp(alpha + A + B + A:B + C )
M1 <- glm(Cirratulidae ~ OrganicM * fFishing + fPeriod,
          data = Benthos,
          family = poisson)

#Exactly the same model
M1 <- glm(Cirratulidae ~ OrganicM + fFishing +
                         OrganicM : fFishing +
                         fPeriod,
          data = Benthos,
          family = poisson)

#Fit the model
#What is the model that we are fitting?
# Cirratulidae_i ~ P(mu_i)
# E(Cirratulidae_i) = mu_i   and   var(Cirratulidae_i) = mu_i
#
#log(mu_i) = alpha + OrganicM_i + fFishing_i +
#            Period_i + OrganicM_i x fFishing_i
```

```{r}
M1 %>% stargazer(type='html')
```

```{r}
#Is the model overdispersed?
E1 <- resid(M1, type = "pearson")
N  <- nrow(Benthos)
p  <- length(coef(M1)) 
sum(E1^2) / (N - p)
#[1] 17.12971
#overdispersed

#or
Overdispersion <- sum(E1^2) / M1$df.res
Overdispersion

#Why do you have overdispersion
#A. Outliers                  => Remove them..but subjective
#B. Missing covariates        => Add them
#C. Missing interactions      => Add them (coplot)
#D. Zero inflation            => ZIP/ZINB
#E. Dependency                => GLMM
#F. Non-linear relationships  => GAM
#G. Wrong link function       => Change it
#H. Variance is bigger than mean => NB

plot(y = E1, x = Benthos$OrganicM)
100 * sum(Benthos$Cirratulidae == 0)/ nrow(Benthos)

#Apply NB GLM
library(MASS)
M2 <- glm.nb(Cirratulidae ~ OrganicM + fFishing +
                         OrganicM : fFishing +
                         fPeriod,
             data = Benthos)

#by default it is the log link

E2 <- resid(M2, type = "pearson")
Overdispersion <- sum(E2^2) / M2$df.res
Overdispersion

#If not lucky....overdispersion > 1.5....
#then:
#Why do you have overdispersion
 #A. Outliers                  => Remove them..but subjective
 #B. Missing covariates        => Add them
 #C. Missing interactions      => Add them (coplot)
 #D. Zero inflation            => ZIP/ZINB
 #E. Dependency                => GLMM
 #F. Non-linear relationships  => GAM
 #G. Wrong link function       => Change it





#What is the model that we are fitting?

# Cirratulidae_i ~ NB(mu_i, k)
# E(Cirratulidae_i) = mu_i
# var(Cirratulidae_i) = mu_i + mu_i^2 / k

#log(mu_i) = alpha + OrganicM_i + fFishing_i +
#            Period_i + OrganicM_i x fFishing_i


print(summary(M2), digits = 2, signif.stars=FALSE)
drop1(M2, test = "Chi")

#the interaction is no longer significant

M3 <- glm.nb(Cirratulidae ~ OrganicM + fFishing+
                         fPeriod, link = "log",
             data =Benthos)

print(summary(M3), digits = 2, signif.stars=FALSE)
drop1(M3, test = "Chi")


#organic material is not significant
M4 <- glm.nb(Cirratulidae ~ fFishing +
                         fPeriod, link = "log",
             data =Benthos)

print(summary(M4), digits = 2, signif.stars=FALSE)
drop1(M4, test = "Chi")

#both fFishing and fPeriod are significant

#Coefficients:
#                Estimate Std. Error z value Pr(>|z|)
#(Intercept)         3.34       0.17    20.1   <2e-16
#fFishingFishing    -0.56       0.15    -3.8    1e-04
#fPeriod2            0.81       0.19     4.2    3e-05
#fPeriod3            0.82       0.19     4.3    2e-05

#NEXT:
#    Model validation
#    Model interpretation
###############################################


#Model validation
par(mfrow = c(2,2))
plot(M4)

#Or:
E4 <- resid(M4, type = "pearson")
F4 <- fitted(M4)
plot(x = F4, 
     y = E4,
     xlab = "Fitted values",
     ylab = "Residuals")

#Influential observations
plot(M4, which = c(4))

#or
par(mfrow=c(1,1))
plot(cooks.distance(M4),
     type = "h",
     ylim=c(0,1))
abline(h=1)



#residuals vs covariates
par(mfrow=c(1,2))
boxplot(E4 ~ fPeriod, data = Benthos)
boxplot(E4 ~ fFishing, data = Benthos)


#Model interpretation
M4 %>% stargazer(type='html')

Model interpretation:
 Model:
 Cirratulidae_i ~ NB(mu_i, 2.471)
 E(Cirratulidae_i) = mu_i
 var(Cirratulidae_i) = mu_i + mu_i^2 / 2.471

 Period 1 & NO_Fishing
 log(mu_i) = 3.34

Period 2& NO_Fishing
 log(mu_i) = 3.34 +0.81

Period 3& NO_Fishing
 log(mu_i) = 3.34 +0.83

 Period 1& Fishing
 log(mu_i) = 3.34 - 0.56

Period 2& Fishing
 log(mu_i) = 3.34 - 0.56 + 0.81

Period 3& Fishing
 log(mu_i) = 3.34 - 0.56 + 0.83
```


#Sketch model fit

```{r eval=FALSE, echo=FALSE}
MyData <- expand.grid(fPeriod = levels(Benthos$fPeriod),
                      fFishing = levels(Benthos$fFishing))
MyData

P <- predict(M4, newdata = MyData,
                    type = "link", se = TRUE)

MyData$Fit   <- exp(P$fit) #need to do exponential
MyData$SeUp  <- exp(P$fit + 2 * P$se.fit)
MyData$SeLow <- exp(P$fit - 2 * P$se.fit)
```
